---
title: "Tweets Stocks Analysis Daily/Hourly"
output: html_document
---

```{r , echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(lubridate)
library(stringr)
library(ggplot2)
library(gridExtra)
library(broom)
library(ggrepel)
# Disable logging of all warning messages
options(warn=-1)
```

# Methodology for extraction of Data
 For analyzing the effect of tweets with the Stock market, we used Twitter streaming api to capture the monitored stocks in realtime and store them onto a file. The raw data in files was processed using the file twitter-data-cleaner.R. This script can be run incrementally and takes care of merging new data with the already set of processed tweets. The script twitter-data-cleaner.R creates a .Rdata file "twitter.RData" that has 4 datasets.
 (TODO add links and information about datasets).
There is a script to process the tweets further and calculate the sentimenet score for each tweet ( tweeter-score-compute.R ), It outputes 4 datasets into a file processed-tweets.RData.  

```{r}
# Following is the list of all the stocks which we have downloaded the tweets and the stocks from Yahoo finance
tickers_symbols <- c("GILD","EIX","GS","AMZN", "RKUS","AAPL","GRPN","XIV","YHOO","VA","MSFT","TSLA","BSX","NVDA","ORCL","EW","CPGX","MRK","V","BXLT","FOXA","ERIC","AVP","TWX","CMCSA","XRX","WY","GNCA","WBA","MO","MA","FOLD","TLT","SNY","RTN","UTX","LOW","MAS","GPT","RICE","IBM","KHC","CDNS","ANTM","HD","INO","OCLR","LULU","SABR","DYN","AXLL","WEN","COH","GOOG","FB","TWTR","XOM","PSX","VLO","PGR","CINF","FAF","JBLU","DAL","HA","ACN","INFY","CTSH")
# Sectors associate with each stock
sectors <- c("Healthcare","Utilities","Financial","Services","Technology","Consumer Goods","Technology","Financial","Technology","Services","Technology","Consumer Goods","Healthcare","Technology","Technology","Healthcare","Basic Materials","Healthcare","Financial","Healthcare","Services","Telecommunications","Consumer Goods","Services","Services","Technology","Industrial Goods","Healthcare","Services","Consumer Goods","Financial","Healthcare",
             "Financial","Healthcare","Industrial Goods","Industrial Goods","Services","Industrial Goods","Financial","Basic Materials","Technology","Consumer Goods","Technology","Healthcare","Services","Healthcare","Technology","Consumer Goods","Technology","Utilities","Basic Materials","Services","Consumer Goods","Technology","Technology","Technology","Basic Materials","Basic Materials","Basic Materials","Financial","Financial","Financial",
             "Services-Airlines","Services-Airlines","Services-Airlines","Technology","Technology","Technology")

selectively_analyzed_symbols <- c("GS","IBM","EIX")
ticker_sector <- data.frame(symbol = tickers_symbols, sector = sectors)
```

Load the Processed tweets and stocks data from github
```{r tweets}
#load(url("https://github.com/goodwillyoga/E107project/raw/master/pulkit/processed-tweets.RData"))
load("/code/CSCIE-107/E107project/pulkit/processed-tweets.RData")
```
```{r stocks}
#load(url("https://github.com/goodwillyoga/E107project/raw/master/pulkit/processed-stocks.RData"))
load("/code/CSCIE-107/E107project/pulkit/processed-stocks.RData")
```
```{r}
# Create helper functions
normalize<-function(m){
  (m - mean(m))/sd(m)
}
```
Let us start looking at distribution of tweets per symbol
```{r,echo=FALSE, message=FALSE}
tweets_est %>% group_by(symbol) %>% 
  summarise(count = n()) %>% inner_join(ticker_sector) %>%
  ggplot(aes(symbol, log(count), color=sector)) + geom_point() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle('Distribution of Tweets per Symbol') + xlab('Symbols')+ylab('Count(log scale) ')+
  theme(legend.position="bottom")
```

Look at distribution of tweets perr day
```{r,echo=FALSE, message=FALSE}
tweets_est %>% group_by(dt) %>% 
  summarise(count = n()) %>%
  ggplot(aes(dt, log(count))) + geom_point() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  theme(legend.position="bottom")+
  ggtitle('Distribution of Tweets per Day') + xlab('Dates')+ylab('Count(log scale) ')
```

Let us look at daily closing prices/Volumes for the 3 choosen stocks
```{r,echo=FALSE, message=FALSE}
dailyStockData %>% filter(symbol %in% selectively_analyzed_symbols) %>% 
  ggplot(aes(day, price)) + geom_line(aes(color=symbol)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle('Varition of Daily closing Price') + xlab('Dates')+ylab('Closing Price ') + facet_wrap(~symbol,scales = "free")
dailyStockData %>% filter(symbol %in% selectively_analyzed_symbols) %>% 
  ggplot(aes(day, log(volume))) + geom_line(aes(color=symbol)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  theme(legend.position="bottom")+
  ggtitle('Varition of Volume') + xlab('Dates')+ylab('Daily Volume ') + facet_wrap(~symbol,scales = "free")
```
```{r,echo=FALSE, message=FALSE}
#We know that Stock markets are closed on weekends, let us compute the trading days so that we can focus on the trading days only
tradingDays <- stocks_est %>% mutate(day = mdy(lastTradeDate)) %>% ungroup() %>% select(day) %>% distinct(day)
```
We will start with modeling of daily stock/tweets data and try to do a linear regression to find relation between volume, closing Price and Daily Price Change with number of tweets. We will represent this once for all the shares and then selectively for 3 choosen stocks to see if we can find if this model 


```{r,echo=FALSE, message=FALSE}
# Normalized tweets count data is stored in daily_tweet_score, Daily Stock data is present in dailyStockData, 
#let us normalize the volume, closing price and Daily Price change and then join it daily_tweet_score
dailyStockTweetNormalized <- dailyStockData %>% 
  group_by(symbol) %>% 
  mutate_each(funs(normalize), volume) %>% 
  mutate_each(funs(normalize),price) %>% 
  mutate_each(funs(normalize),prcChange) %>% 
  inner_join(daily_tweet_score, by=c("symbol"="symbol", "day"="dt")) %>% select(symbol,price,volume,open,sector,date_timelb,day,prcChange,count,avgScore)

fits <- dailyStockTweetNormalized %>%
  group_by(symbol) %>%
  do(mod = lm(volume ~ count, data = .))

results_volume_count <- tidy(fits,mod, conf.int = TRUE) %>% filter(term=='count') %>% inner_join(ticker_sector)
colnames(results_volume_count)[6]<- 'pval'

# Plot for all stocks
ggplot(results_volume_count, aes(symbol, pval, color=sector)) + geom_point() + geom_hline(yintercept = .05, color = 'red') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("volume~count model for all Stocks") +
  geom_text_repel(aes(symbol, pval, label=symbol), data = filter(results_volume_count,  pval <.05 )) +
  theme(legend.position="bottom")

# Let us model closing price vs count for each stock
fits <- dailyStockTweetNormalized %>%
  group_by(symbol) %>%
  do(mod = lm(price ~ count, data = .))

results_price_count <- tidy(fits,mod, conf.int = TRUE) %>% filter(term=='count') %>% inner_join(ticker_sector)
colnames(results_price_count)[6]<- 'pval'

# Plot for all stocks
ggplot(results_price_count, aes(symbol, pval, color=sector)) + geom_point() + geom_hline(yintercept = .05, color = 'red') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Closing Price~count model for all Stocks") +
  geom_text_repel(aes(symbol, pval, label=symbol), data = filter(results_price_count,  pval <.05 )) +
  theme(legend.position="bottom")


# Let us model Price Change vs count for all stocks
fits <- dailyStockTweetNormalized %>%
  group_by(symbol) %>%
  do(mod = lm(prcChange ~ count, data = .))

results_prcChange_count <- tidy(fits,mod, conf.int = TRUE) %>% filter(term=='count') %>% inner_join(ticker_sector)
colnames(results_prcChange_count)[6]<- 'pval'

# Plot for all stocks
ggplot(results_prcChange_count, aes(symbol, pval, color=sector)) + geom_point() + geom_hline(yintercept = .05, color = 'red') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Daily Price Change~count model for all Stocks") +
  geom_text_repel(aes(symbol, pval, label=symbol), data = filter(results_prcChange_count,  pval <.05 )) +
  theme(legend.position="bottom")

# Let us plot specific for the 3 stocks 
p1<- results_volume_count %>% filter(symbol %in% selectively_analyzed_symbols) %>% ggplot( aes(symbol, pval, color=symbol)) + geom_point() + geom_hline(yintercept = .05, color = 'red') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("volume~count model")+ theme(legend.position="bottom")

p2 <- results_price_count %>% filter(symbol %in% selectively_analyzed_symbols)  %>% ggplot( aes(symbol,pval, color=symbol))+geom_point()+geom_hline(yintercept = .05, color='red')+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+ ggtitle("Closing Price~count model")+ theme(legend.position="bottom")

p3<- results_prcChange_count %>% filter(symbol %in% selectively_analyzed_symbols)  %>% ggplot( aes(symbol,pval, color=symbol))+geom_point()+geom_hline(yintercept = .05, color='red')+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+ ggtitle("Price Change~count model")+ theme(legend.position="bottom")

grid.arrange(arrangeGrob(p1+ 
                           theme(legend.position="none"), p2+ 
                           theme(legend.position="none"), p3+ 
                           theme(legend.position="none"), nrow=1), top = "P-Value for standardized")

# Let us plot the confidence intervals
p1 <- results_volume_count %>% 
  filter(symbol %in% selectively_analyzed_symbols) %>% 
  select(symbol,estimate, pval,`conf.low`, `conf.high`) %>% 
  ggplot( aes(x = symbol, y = estimate, color = symbol)) +
  geom_point(size = 4) + ggtitle("volume~count model")+
  geom_errorbar(aes(ymax = `conf.high`, ymin = `conf.low`))

p2 <-results_price_count %>% 
  filter(symbol %in% selectively_analyzed_symbols) %>% 
  select(symbol,estimate, pval,`conf.low`, `conf.high`) %>% 
  ggplot( aes(x = symbol, y = estimate, color = symbol)) +
  geom_point(size = 4) + ggtitle("Closing Price~count model")+
  geom_errorbar(aes(ymax = `conf.high`, ymin = `conf.low`))

p3 <-results_prcChange_count %>% 
  filter(symbol %in% selectively_analyzed_symbols) %>% 
  select(symbol,estimate, pval,`conf.low`, `conf.high`) %>% 
  ggplot( aes(x = symbol, y = estimate, color = symbol)) +
  geom_point(size = 4) + ggtitle("Price Change~count model")+
  geom_errorbar(aes(ymax = `conf.high`, ymin = `conf.low`))

grid.arrange(arrangeGrob(p1+ 
                           theme(legend.position="none"), p2+ 
                           theme(legend.position="none"), p3+ 
                           theme(legend.position="none"), nrow=1), top = "Confidence Interval Plots")

# Cleanup all the objects created in this block
rm(p1, p2, p3, results_prcChange_count, results_price_count, results_volume_count,fits)

```

We can see that for GS we can have a successfull model of 


```{r ,echo=FALSE, message=FALSE}
# Let us try to model now based on the sentiment scores
fits <- dailyStockTweetNormalized %>%
  group_by(symbol) %>%
  do(mod = lm(volume ~ avgScore, data = .))

results_volume <- tidy(fits,mod, conf.int = TRUE) %>% filter(term=='avgScore') %>% inner_join(ticker_sector)
colnames(results_volume)[6]<- 'pval'

# Plot for all stocks
ggplot(results_volume, aes(symbol, pval, color=sector)) + geom_point() + geom_hline(yintercept = .05, color = 'red') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("volume~avgScore model") +
  geom_text_repel(aes(symbol, pval, label=symbol), data = filter(results_volume,  pval <.05 )) +
  theme(legend.position="bottom")

# Let us model closing price vs count for each stock
fits <- dailyStockTweetNormalized %>%
  group_by(symbol) %>%
  do(mod = lm(price ~ avgScore, data = .))

results_price <- tidy(fits,mod, conf.int = TRUE) %>% filter(term=='avgScore') %>% inner_join(ticker_sector)
colnames(results_price)[6]<- 'pval'

# Plot for all stocks
ggplot(results_price, aes(symbol, pval, color=sector)) + geom_point() + geom_hline(yintercept = .05, color = 'red') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Closing Price~avgScore model") +
  geom_text_repel(aes(symbol, pval, label=symbol), data = filter(results_price,  pval <.05 )) +
  theme(legend.position="bottom")


# Let us model Price Change vs count for all stocks
fits <- dailyStockTweetNormalized %>%
  group_by(symbol) %>%
  do(mod = lm(prcChange ~ avgScore, data = .))

results_prcChange <- tidy(fits,mod, conf.int = TRUE) %>% filter(term=='avgScore') %>% inner_join(ticker_sector)
colnames(results_prcChange)[6]<- 'pval'

# Plot for all stocks
ggplot(results_prcChange, aes(symbol, pval, color=sector)) + geom_point() + geom_hline(yintercept = .05, color = 'red') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Daily Price Change~avgScore model") +
  geom_text_repel(aes(symbol, pval, label=symbol), data = filter(results_prcChange,  pval <.05 )) +
  theme(legend.position="bottom")

# Let us plot specific for the 3 stocks 
p1<- results_volume %>% filter(symbol %in% selectively_analyzed_symbols) %>% ggplot( aes(symbol, pval, color=symbol)) + geom_point() + geom_hline(yintercept = .05, color = 'red') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("volume~avgScore model")+ theme(legend.position="bottom")

p2 <- results_price %>% filter(symbol %in% selectively_analyzed_symbols)  %>% ggplot( aes(symbol,pval, color=symbol))+geom_point()+geom_hline(yintercept = .05, color='red')+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+ ggtitle("Closing Price~avgScore model")+ theme(legend.position="bottom")

p3<- results_prcChange %>% filter(symbol %in% selectively_analyzed_symbols)  %>% ggplot( aes(symbol,pval, color=symbol))+geom_point()+geom_hline(yintercept = .05, color='red')+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+ ggtitle("Price Change~avgScore model")+ theme(legend.position="bottom")

grid.arrange(arrangeGrob(p1+ 
                           theme(legend.position="none"), p2+ 
                           theme(legend.position="none"), p3+ 
                           theme(legend.position="none"), nrow=1), top = "P-Value for standardized")

# Let us plot the confidence intervals
p1 <- results_volume %>% 
  filter(symbol %in% selectively_analyzed_symbols) %>% 
  select(symbol,estimate, pval,`conf.low`, `conf.high`) %>% 
  ggplot( aes(x = symbol, y = estimate, color = symbol)) +
  geom_point(size = 4) + ggtitle("volume~avgScore model")+
  geom_errorbar(aes(ymax = `conf.high`, ymin = `conf.low`))

p2 <-results_price %>% 
  filter(symbol %in% selectively_analyzed_symbols) %>% 
  select(symbol,estimate, pval,`conf.low`, `conf.high`) %>% 
  ggplot( aes(x = symbol, y = estimate, color = symbol)) +
  geom_point(size = 4) + ggtitle("Closing Price~avgScore model")+
  geom_errorbar(aes(ymax = `conf.high`, ymin = `conf.low`))

p3 <-results_prcChange %>% 
  filter(symbol %in% selectively_analyzed_symbols) %>% 
  select(symbol,estimate, pval,`conf.low`, `conf.high`) %>% 
  ggplot( aes(x = symbol, y = estimate, color = symbol)) +
  geom_point(size = 4) + ggtitle("Price Change~avgScore model")+
  geom_errorbar(aes(ymax = `conf.high`, ymin = `conf.low`))

grid.arrange(arrangeGrob(p1+ 
                           theme(legend.position="none"), p2+ 
                           theme(legend.position="none"), p3+ 
                           theme(legend.position="none"), nrow=1), top = "Confidence Interval Plots")

rm(p1, p2, p3, results_prcChange, results_price, results_volume,fits)

```